Assessing the reliability of an online behavioural laterality battery: A pre-registered study.

### 2020-11-27 online_laterality_reliability.Rmd
revised manuscript

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. Online testing can allow for much larger sample sizes but limits the amount of experimental control that is feasible. Here we considered whether online testing could give valid and reliable results on four tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task. We recruited 392 participants, oversampling left-handers, who completed the battery twice. Three of the tasks showed evidence of both validity and reliability, insofar as they showed hemispheric advantages in the expected direction and test-retest reliability of at least *r*= .75. The reliability of the rhyme decision task was less satisfactory (*r*= .63). We also found that extreme left-handers were more likely to depart from typical lateralisation. Lateralisation across the two language tasks (dichotic listening and rhyme judgement) was weakly correlated, but unrelated to lateralisation on the chimeric faces task. We conclude that three of the tasks, dichotic listening, chimeric faces and finger tapping, show considerable promise for online evaluation of cerebral lateralisation. The results do not support the notion of a common underlying cause of lateralisation for language and face processing.

@@

Behavioural laterality tasks have long been used to establish hemispheric dominance for certain cognitive functions. Results from visual half-field and dichotic listening paradigms support the view that the left hemisphere plays the dominant role in language processing (Hugdahl, 2000; Ocklenburg & Güntürkün, 2018). By contrast, visual half-field paradigms indicate that the right hemisphere is involved in the processing of faces, global feature processing, and visuo-spatial attention (Brederoo, Nieuwenstein, Cornelissen, & Lorist, 2019). Below, we review how each paradigm has been implemented to study language lateralisation and the detection of emotion in faces, and what is known about reliability and validity.

@@

Lateralised visual word processing has been studied using the visual half-field paradigm. In this paradigm, visual stimuli are briefly presented to the left or right of a central fixation stimulus. The visual pathways from the left and right sides of space to the visual cortex are crossed for non-foveal viewing, and so the right visual-field (RVF) projects directly to the left hemisphere and the left visual-field (LVF) to the right hemisphere. The corpus callosum allows for transfer between hemispheres, but the rationale of the method is that there will be a processing advantage for language stimuli that are directly projected to the language-dominant hemisphere. This method has been used since the 1950s (Mishkin & Forgays, 1952). Bourne (2006) reviewed the literature, which consistently shows advantageous processing for letter strings presented in the RVF, relative to the LVF, in most people.

@@

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Hunter and Brysbaert argued that this is the result of stimuli competing during bilateral presentation, enhancing the advantage to the stimulus presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation to prevent eye movements. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) used two visual half-field tasks, word naming (192 trials) and picture naming (160 trials) in which stimuli were bilaterally presented and masked at offset. The two tasks correlated strongly (*r*= .8, 95% CI .598-.956) in a sample of 26 left-handers. Encouragingly, the LIs from these visual half-field tasks also correlated positively with laterality from functional Magnetic Resonance Imaging (fMRI) on a word generation task in 10 left-handers. However, the small sample size meant that the confidence intervals around these estimates were very wide: word naming, *r*= .63 (95% CI = .001-.902), picture naming, *r*= .77 (95% CI = .272-.943). More recently, however, Van der Haegen, Cai, Seurinck, and Brysbaert (2011) reported highly similar findings in a somewhat larger sample of 50 participants. Again, participants completed behavioural word and picture naming visual half-field tasks and then a word generation task under fMRI. It is reassuring that the LIs for the behavioural measures were correlated with the fMRI LI: word naming, *r* = .64, picture naming, *r* = .65. Van der Haegen et al. also report a set of analyses involving 250 participants who completed the behavioural visual half-field tasks while their eye movements were monitored. Two sets of LIs were calculated; one controlling for fixation error and one that did not. Importantly, the LIs which did and did not control for fixation error were highly correlated for word naming, *r* = .94, and picture naming,*r* =.98, indicating that the inclusion of eye movement control adds little additional value when a visual half-field task employs mechanisms to encoruage central fixation and bilateral presentation.

A promising line of work has emerged which further highlights the potential for well designed visual half-field tasks to assess language lateralisation. First introduced by Willemin et al. (2016), the translingual lexical decision task involves participants viewing letter stings either side of fixation and judging whether the left, right, or neither of the letter strings was a valid word. With 100 participants completing 256 trials, Willemin et al. reported a RVF advantage at the population level. Since then this effect has been replicated in a sample of 496 participants covering six international languages, with evidence of a RVF for each language studied. Not only does this task show validity, it seems reliable given Brederoo et al.'s (2020) observation that the task yields good split-half reliability when tested with 122 participants, *r* = 0.86 for error rates and *r* = 0.80 for response times.

@@

Simultaneous presentation of different auditory stimuli to left and right ears for immediate recall was a method originally used in studies of attention, but it became evident that there was usually an advantage for verbal stimuli presented to the right ear (Kimura, 1967). Both ears project to both hemispheres, but the weaker ipsilateral pathways appear to be suppressed when the two sides compete, and so the right ear, with a contralateral projection to the left hemisphere, will have an advantage in tasks that involve language processing.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed optimally, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explained their justification for each. Their recommendations were: (1) stop consonant-vowel (CV) syllables as stimulus materials; (2) CV stimulus pairs from the same voicing category; (3) alternating trials of voiced and unvoiced stimulus pairs; (4) free-recall instruction; (5) one stimulus pair per trial; (6) a paradigm length of 120 dichotic trials; (7) all stimulus pairs presented at equal frequency to the two sides; and (8) inclusion of binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded test-retest reliability of *r*= .93 for verbal responses and *r*= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

To date, little work has been conducted online with visual stimuli. By contrast, the dichotic listening paradigm has been implemented online with considerable success. Bless et al. (2013) developed a mobile device version of the consonant-vowel dichotic listening paradigm, with 36 simultaneous sound pairs. Under laboratory conditions, with a sample of 76 (33 Norwegian, 43 Austrialian) participants, they reported a test-retest reliability of *r*= 0.78 when using the mobile app. Furthermore, both subgroups of participants showed clear right ear advantages at the population level. In a subsequent large-scale study, Bless et al. (2015) administered the same application to a sample of over 4,000 participants with more than 60 different language backgrounds. Again Bless et al. (2015) reported a clear right ear advanatge at the population-level. They then used the large dataset to examine cultural/lingusitic effects on the degree on lateralisation; reporting that the degree of lateralization appears to be modulated by linguistic background. Together these studies demonstrate the potential for online laterality research to generate large amounts of data while maintaining a tolerable degree of reliability.

@@

With regards to visual half-field tasks for studying language lateralisation, there is evidence that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Weekes, Capetillo-Cunliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well-powered study involving 100 participants, conducted by Willemin et al. (2016), reported no hint of any effect of handedness, when coded as left and right, or when divided into left (N = 19), mixed (N = 24) and right-handed (N = 57) groups, on the visual field advantage during a lexical decision task. This result illustrates the recurring challenge of interpreting null results in this field: do they reflect a true lack of relationship, inadequate power to detect small effects, or specific methodological factors? Brederoo et al. (2020) reported good split-half reliability for this task, suggesting the lack of handedness effect is unlikely to be due solely to measurement error. We may note that Willemin et al. asked participants to respond via manual response, whereas other studies have involved an active speech production component; manual responding could possibly interact with handedness to influence the laterality index on this task.

On dichotic listening tasks, the clear right ear (i.e. left hemisphere) advantage that is noted at the population level (see Hugdahl et al., 2009, and Westerhausen, 2019, for a review) is reduced in left-handers (Bethmann, Tempelmann, De Bleser, Scheich, & Brechmann, 2007; Bryden, 1970, 1988; Hugdahl, 2003). More recently, Karlsson, Johnstone, and Carey (2019) employed a consonant-vowel version of the dichotic listening task where participants reported which one of six syllables that they heard most clearly. In study 1, 87% of right-handers and 78% of non-right-handers showed a right ear advantage. For study 2, 85% of right-handers and 79% of non-right-handers showed a right ear advantage.

For chimeric faces tasks, which typically show a left visual field advantage, there again appears to be a greater departure from the typical pattern of laterality for left-handers (e.g. Gilbert & Bakan, 1973; Levy et al., 1983; Rozkowski & Snelbecker, 1982). This was also studied by Karlsson et al. (2019). When viewing a single face with a different emotion shown in either hemiface, 75% of right-handers and 60% of non-right-handers showed a leftwards bias. For the second chimeric faces task that involved two chimeric faces arranged vertically, a similar pattern emerged. 73% of right-handers and 57% of non-right-handers showed a leftwards bias.

@@

To date the focus has been in testing whether lateralisation shows  complementarity across tasks that are usually mediated by left and right hemispheres, especially whether there is a relationship between language vs visuospatial functioning. However, emerging lines of evidence have suggested that even within domains there may be dissociated patterns of lateralisation across similar tasks. Woodhead, Bradshaw, Wilson, Thompson, and Bishop (2019) reported data for 37 participants engaged in six language tasks under transcranial Doppler sonography (fTCD) and examined the factorial structure of lateralisation across tasks using structural equation modeling. They pitted two models against each other. The first assumed that language laterality is a unidimensional trait and a single factor was used to model covariance. The second assumed that that two laterality factors underlie the pattern of covariance. The results supported a two factor model at the population level. However, further examination of the data revealed that this was driven by seven left-handers. Woodhead, Thompson, Karlsson, and Bishop (2020) examined this factor structure in an extended sample of 74 participants with 43 right-handers and 31 left-handers. The data this time indicated that a two factor model was preferred for both right- and left-handers, with the preference for a two factor model being stronger for left-handers. Together, this indicated that even with the domain of language laterality indices may be dissociated.

@@

When selecting specific tasks in our online battery we were strongly guided by pragmatic considerations, i.e. how feasible would it be to administer the task remotely via an online platform. We aimed to include at least two language measures, one tapping into receptive language and one expressive, in order to test the notion that language laterality can dissociate across different tasks. We also aimed to include a task that involved face processing so that we could test for complementarity of lateralisation across tasks that usually lateralise to left and right hemispheres. Finally, a behavioural measure of motor laterality was included, to potentially provide an online analogue of relative hand skill similar to that used by Annett (1970).

@@

Given that numerous studies looking at the relationship lateralisation on language and non-language tasks have focused on the use of face stimuli (e.g. Brederoo et al., 2020), we chose to include a face task to conduct similar comparisons. Regarding the chimeric faces, Levy et al. (1983) reported good split-half reliability while Karlsson et al. had reported that, like the dichotic listening task, the the chimeric faces task is sensitive to the differences in perceptual asymmetry in right- and non-right-handers.

Our final online measure was a finger-tapping test used as an online alternative to the well-known Annett (1970) pegboard task. There is suggestive evidence that a continuous measure of relative skill of the two hands may be more effective than more conventional hand preference measures for capturing individual differences in motor laterality that relate to language lateralisation (Annett, 2002; Flowers & Hudson, 2013). Performance measures such as this can provide a continuous measure of subtle variation in relative hand skill, and also correlate well with measures of hand preference (e.g. Steenhuis & Bryden, 1999). The pegboard task is one of the most widely used performance tasks (see Hodgson & Hudson, 2016, for discussion), but the requirement for apparatus makes it unfeasible as an online test. We therefore took the opportunity to explore the reliability and validity of a finger-tapping task that required the participant to repeatedly tap a short sequence on the computer keyboard with the index finger of the preferred or nonpreferred hand, which capture one aspect of pegmoving, namely precise aiming of sequential movements.

@@

Our final prediction (hypothesis 3) related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult to gather large amounts of data to establish test-retest reliability. For instance, van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of the individual measures. In the pre-registration for our current study, we predicted that for right-handers, a model where the two language tasks are correlated with each other but not with the chimeric faces (model A) would be the best fit to the data, compared to either a model in which laterality indices are correlated between the language and chimeric face tasks (model B), or a base model where each task is independently lateralised (model C). Given our previous findings using fTCD to study dissociations in language laterality (Woodhead, Bradshaw, Wilson, Thompson, & Bishop, 2019), we further predicted that model fit may differ for right- and left-handers, with more dissociation between tasks in the latter group.

@@

As our ultimate goal is to use these measures in individual differences research, we aimed to optimise the task to achieve observed test-retest reliability of at least *r*= .65. Our initial power analysis showed that to achieve sufficient precision around the estimate of reliability, we would require 300 participants. Out of the 300 participants, we originally specified that 80 would be left-handers. This would give us power to apply our planned model-fitting analysis to left-handers as a separate group.  In practice, our final sample exceeded 300, because we discarded participants from individual tasks for failing to meet prespecified criteria (see below).  
In total, 441 native English speakers were initially recruited through Prolific Academic (www.prolific.co) with all these participants completing both sessions,  Prolific is a platform for recruiting and screening paid research participants according to experimenter-specified criteria. Participants were 18- to 50-years-old, had normal or corrected-to-normal vision and reported no history of neurological disease (e.g. head injury, brain tumour, stroke) or significant hearing loss or impairment.

@@

The basic demographics questionnaire asked participants to report age, gender, years in education, and whether they were bilingual. We also asked questions about handedness and footedness. For handedness, we asked: which is your dominant hand (i.e. which hand do you prefer to use for tasks such as writing, cutting, and catching a ball?). For footedness, we asked: which foot do you normally use to step up on a ladder/step? For both questions, participants could answer "left", "right", or "no preference / Don't know".

@@

Adapted versions of the Miles test (Miles, 1929) and the Porta test (Porac & Coren, 1976) were used to determine each participant’s eye dominance in central gaze (i.e., when looking straight ahead). Each test classified participants as being either left or right eye dominant. We did not have specific predictions about ocular dominance, but gathered this information to further characterise behavioural laterality of the sample. In practice, the two tests agreed perfectly, so we report here just the Miles test.

@@

The rhyme decision task was developed for this project as a novel visual half-field technique designed to engage lateralised language processes. On each trial, participants were asked to decide whether a foveally presented letter string rhymed with the name of an image shown in either the LVF or RVF. With online presentation, a particular concern is a potential for eye movements away from fixation. We relied on the fact that participants had to focus initially on the central written word to do the task, and thus the task encouraged fixation to a central location at the start of the trial. We subsequently displayed pictures under brief bilateral presentation, which has been shown to give adequate control for eye movements in previous visual half-field experiments with verbal naming (Beaumont, 1982). Studies that directly monitored eye movements in this setting reported failures of fixation on only 0.5% of trials (Geffen, Bradshaw & Nettleton, 1972; but see also Bourne, 2006). Pictures of familiar objects were presented for rhyme matching to the written word, to ensure that the participant would have to covertly name the picture, thereby engaging left hemisphere speech processing. We used covert rather than overt naming (cf. Hunter & Brysbaert, 2008) to make the task suitable for online presentation. This obviated the need for measuring latencies of oral responses, which would be challenging in the online context, both in terms of need for a high-quality microphone, and possible intrusion of background noises interfering with recording. Furthermore, the evaluation of oral responses would require additional scoring following data collection which would be a taxing procedure for researchers. It would also require additional ethical approvals with regards to the storing and processing of the data.

@@

```{r read_dat, echo- FALSE, include= FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(corrr)
library(flextable)
library(cocor)
options(knitr.kable.NA = '') #should suppress NA in tables
# read in data from csv
#LIs <- read.csv('https://osf.io/x5jd2/download',stringsAsFactors = F)
#We will replace with read from OSF, but better stick with LIs to ensure final version is the one we use 
allsum <- read.csv('./data/allsum.csv',stringsAsFactors = F)

```{r prepdat,echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}
# remove those without day 2 to make LIs file - we will just use this
# reformat NA
allsum$day2.complete <- ifelse(is.na(allsum$day2.complete), 0, allsum$day2.complete)
LIs <- allsum[allsum$day2.complete == 1,]

>>>>

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. Online testing can allow for much larger sample sizes but limits the amount of experimental control that is feasible. Here we considered whether online testing could give valid and reliable results on four tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task. We recruited 392 participants, oversampling left-handers, who completed the battery twice. Three of the tasks showed evidence of both validity and reliability, insofar as they showed hemispheric advantages in the expected direction and test-retest reliability of at least *r*= .75. The reliability of the rhyme decision task was less satisfactory (*r*= .62). We also confirmed a prediction that extreme left-handers were more likely to depart from typical lateralisation. Lateralisation across the two language tasks (dichotic listening and rhyme judgement) was weakly correlated, but unrelated to lateralisation on the chimeric faces task. We conclude that three of the tasks, dichotic listening, chimeric faces and finger tapping, show considerable promise for online evaluation of cerebral lateralisation. The results supported a common language factor that influenced language tasks but found no evidence for a common underlying cause of lateralisation for language and face processing.

@@

Behavioural laterality tasks have long been used to establish hemispheric dominance for certain cognitive functions. Results from visual half-field and dichotic listening paradigms support the view that the left hemisphere plays the dominant role in language processing (Hugdahl, 2000; Ocklenburg & Güntürkün, 2018). By contrast, visual half-field paradigms indicate that the right hemisphere is involved in the processing of faces, global feature processing, and visuospatial attention (Brederoo, Nieuwenstein, Cornelissen, & Lorist, 2019). Below, we review how each paradigm has been implemented to study language lateralisation and the detection of emotion in faces, and what is known about reliability and validity.

@@

Lateralised visual word processing has been studied using the visual half-field paradigm. In this paradigm, visual stimuli are briefly presented to the left or right of a central fixation stimulus. The visual pathways from the left and right sides of space to the visual cortex are crossed for non-foveal viewing, and so the right visual-field (RVF) projects directly to the left hemisphere and the left visual-field (LVF) to the right hemisphere. The corpus callosum allows for transfer between hemispheres, but the rationale of the method is that there will be a processing advantage for language stimuli that are directly projected to the language-dominant hemisphere. This method has been used since the 1950s (Mishkin & Forgays, 1952). Bourne (2006) reviewed the literature, which consistently shows superior processing for letter strings presented in the RVF, relative to the LVF, in most people.

@@

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is influenced by several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Hunter and Brysbaert argued that this competition between stimuli during bilateral presentation enhances the advantage to the stimulus presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation to prevent eye movements. When these methodological aspects are addressed, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) used two visual half-field tasks, word naming (192 trials) and picture naming (160 trials) in which stimuli were bilaterally presented and masked at offset. The two tasks correlated strongly (*r*= .80, *95% CI*= .598-.956) in a sample of 26 left-handers. Encouragingly, the LIs from these visual half-field tasks also correlated positively with laterality from functional Magnetic Resonance Imaging (fMRI) on a word generation task in 10 left-handers. However, the small sample size meant that the confidence intervals around these estimates were very wide: word naming, *r*= .63 (*95% CI*= .001-.902), picture naming, *r*= .77 (*95% CI*= .272-.943). More recently, however, Van der Haegen, Cai, Seurinck, and Brysbaert (2011) reported similar findings in a sample of 50 participants. Again, participants completed behavioural word and picture naming visual half-field tasks and then a word generation task under fMRI. It is reassuring that the LIs for the behavioural measures were correlated with the fMRI LI: word naming, *r* = .64, picture naming, *r* = .65. Van der Haegen et al. also report a set of analyses involving 250 participants who completed the behavioural visual half-field tasks while their eye movements were monitored. Two sets of LIs were calculated; one that controlled for fixation error and one that did not. Importantly, the LIs which did and did not control for fixation error were highly correlated with word naming, *r* = .94, and picture naming,*r* =.98, indicating that the inclusion of eye movement control adds little additional value when a visual half-field task employs mechanisms to encourage central fixation and bilateral presentation.

A promising line of work has emerged which further highlights the potential for well-designed visual half-field tasks to assess language lateralisation. First introduced by Willemin et al. (2016), the translingual lexical decision task involves participants viewing letter stings either side of fixation and judging whether the left, right, or neither of the letter strings is a valid word. With 100 participants completing 256 trials, Willemin et al. reported a RVF advantage at the population level. Since then this effect has been replicated in a sample of 496 participants covering six international languages, with evidence of a RVF for each language studied (Hausmann et al., 2019). Not only does this task show validity, it seems reliable given Brederoo et al.'s (2020) observation that the task yields good split-half reliability when tested with 122 participants: *r* = 0.86 for error rates and *r* = 0.80 for response times.

@@

Simultaneous presentation of different auditory stimuli to left and right ears for immediate recall was a method originally used in studies of attention, but it became evident that there was usually better recall of verbal stimuli presented to the right ear (Kimura, 1967). Both ears project to both hemispheres, but the weaker ipsilateral pathways appear to be suppressed when the two sides compete, and so the right ear, with a contralateral projection to the left hemisphere, will have an advantage in tasks that involve language processing.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed optimally, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explained their justification for each. Their recommendations were: (1) stop consonant-vowel (CV) syllables as stimulus materials; (2) CV stimulus pairs from the same voicing category; (3) alternating trials of voiced and unvoiced stimulus pairs; (4) free-recall instruction; (5) one stimulus pair per trial; (6) a paradigm length of 120 dichotic trials; (7) all stimulus pairs presented at an equal frequency to the two sides; and (8) inclusion of binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded test-retest reliability of *r*= .93 for verbal responses and *r*= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

To date, little work has been conducted online with visual stimuli. By contrast, the dichotic listening paradigm has been implemented online with considerable success. Bless et al. (2013) developed a mobile device version of the consonant-vowel dichotic listening paradigm, with 36 simultaneous sound pairs. Under laboratory conditions, with a sample of 76 (33 Norwegian, 43 Australian) participants, they reported test-retest reliability of *r*= 0.78 when using the mobile app. Furthermore, both subgroups of participants showed clear right ear advantages at the population level. In a subsequent large-scale study, Bless et al. (2015) administered the same application to a sample of over 4,000 participants with more than 60 different language backgrounds. Again Bless et al. (2015) reported a clear right ear advantage at the population-level. They then used the large dataset to examine cultural/linguistic effects on the degree on lateralisation; reporting that the degree of lateralization appears to be modulated by linguistic background. Together these studies demonstrate the potential for online laterality research to generate large amounts of data while maintaining a tolerable degree of reliability.

@@

With regards to visual half-field tasks for studying language lateralisation, there is evidence that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Weekes, Capetillo-Cunliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well-powered study involving 100 participants, conducted by Willemin et al. (2016), reported no hint of any effect of handedness, when coded as left and right, or when divided into left (N = 19), mixed (N = 24) and right-handed (N = 57) groups, on the visual field advantage during a lexical decision task. This result illustrates the recurring challenge of interpreting null results in this field: do they reflect a true lack of relationship, inadequate power to detect small effects, or specific methodological factors? Brederoo et al. (2020) reported good split-half reliability for this task, suggesting the lack of handedness effect is unlikely to be due solely to measurement error. We may note that Willemin et al. asked participants to respond via manual response, whereas other studies have involved an active speech production component; manual responding could interact with handedness to influence the laterality index on this task.

On dichotic listening tasks, the clear right ear (i.e. left hemisphere) advantage that is noted at the population level (see Hugdahl et al., 2009, and Westerhausen, 2019, for a review) is reduced in left-handers (Bethmann, Tempelmann, De Bleser, Scheich, & Brechmann, 2007; Bryden, 1970, 1988; Hugdahl, 2003). Recently, Karlsson, Johnstone, and Carey (2019) employed a consonant-vowel version of the dichotic listening task where participants reported which one of six syllables that they heard most clearly. In study 1, 87% of right-handers and 78% of non-right-handers showed a right ear advantage. For study 2, 85% of right-handers and 79% of non-right-handers showed a right ear advantage.

For chimeric faces tasks, which typically show a left visual field advantage, there again appears to be a greater departure from the typical pattern of laterality for left-handers (e.g. Gilbert & Bakan, 1973; Levy et al., 1983; Rozkowski & Snelbecker, 1982). This was also studied by Karlsson et al. (2019). When viewing a single face with a different emotion shown in either hemiface, 75% of right-handers and 60% of non-right-handers showed a leftwards bias. For the second chimeric faces task that involved two chimeric faces arranged vertically, a similar pattern emerged: 73% of right-handers and 57% of non-right-handers showed a leftwards bias.

@@

To date, the focus has been in testing whether lateralisation shows  complementarity across tasks that are usually mediated by the left and right hemispheres, especially whether there is a relationship between language vs visuospatial functioning. However, emerging lines of evidence have suggested that even within domains there may be dissociated patterns of lateralisation across similar tasks. Woodhead, Bradshaw, Wilson, Thompson, and Bishop (2019) reported data for 37 participants engaged in six language tasks under transcranial Doppler sonography (fTCD) and examined the factorial structure of lateralisation across tasks using structural equation modeling. They pitted two models against each other. The first assumed that language laterality is a unidimensional trait and a single factor was used to model covariance. The second assumed that two laterality factors underlie the pattern of covariance. The results supported a two-factor model at the population level. However, further examination of the data revealed that this was driven by seven left-handers. Furthermore, Woodhead, Thompson, Karlsson, and Bishop (2020) examined this factor structure in an extended sample of 74 participants (43 right-handers and 31 left-handers). The data this time indicated that a two-factor model was preferred for both right- and left-handers, with the preference for a two-factor model being stronger for left-handers. Together these data suggest that  laterality indices within the domain of language may be dissociated.

@@

When selecting specific tasks in our online battery we were strongly guided by pragmatic considerations, i.e. how feasible would it be to administer the task remotely via an online platform. We aimed to include at least two language measures, one tapping into receptive language and one expressive, to test the notion that language laterality can dissociate across different tasks (i.e. Woodhead et al., 2019, 2020). We also aimed to include a task that involved face processing so that we could test for complementarity of lateralisation across tasks that usually lateralise to left and right hemispheres. Finally, a behavioural measure of motor laterality was included, to potentially provide an online analogue of relative hand skill similar to that used by Annett (1970).

@@

Given that numerous studies looking at the relationship between lateralisation on language and non-language tasks have focused on the use of face stimuli (e.g. Brederoo et al., 2020), we chose to include a face task to conduct similar comparisons. Regarding the chimeric faces task, Levy et al. (1983) reported good split-half reliability while Karlsson et al. had reported that, like the dichotic listening task, the chimeric faces task is sensitive to the differences in perceptual asymmetry in right- and non-right-handers.

Our final online measure was a finger-tapping test used as an online alternative to the well-known Annett (1970) pegboard task. There is suggestive evidence that a continuous measure of relative skill of the two hands may be more effective than more conventional hand preference measures for capturing individual differences in motor laterality that relate to language lateralisation (Annett, 2002; Flowers & Hudson, 2013). Performance measures such as this can provide a continuous measure of subtle variation in relative hand skill, and also correlate well with measures of hand preference (e.g. Steenhuis & Bryden, 1999). The pegboard task is one of the most widely used performance tasks (see Hodgson & Hudson, 2016, for discussion), but the requirement for apparatus makes it unfeasible as an online test. We therefore took the opportunity to explore the reliability and validity of a finger-tapping task that required the participant to repeatedly tap a short sequence on the computer keyboard with the index finger of the preferred or non-preferred hand, which captures one aspect of peg-moving, namely precise aiming of sequential movements.

@@

Our final prediction (hypothesis 3) related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult to gather large amounts of data to establish test-retest reliability. For instance, van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of the individual measures. In the pre-registration for our current study, we predicted that for right-handers, a two-factor model where the two language tasks are correlated with each other but not with the chimeric faces (model 2F) would be the best fit to the data, compared to either a single factor model in which laterality indices are correlated between the language and chimeric face tasks (model 1F), or a base model where each task is independently lateralised (model 3F). Given our previous findings using fTCD to study dissociations in language laterality (Woodhead et al., 2019, 2020), we further predicted that model fit may differ for right- and left-handers, with more dissociation between tasks in the latter group.  While we report this analysis in the Supplementary Materials, we report an analysis of the whole sample in the main body as we found that the results from subgroup analyses were very unstable with the small sample of left-handers who were eligible for analysis, and could favour different models when quite minor changes were made to the sample.

@@

As our ultimate goal is to use these measures in individual differences research, we aimed to optimise the task to achieve observed test-retest reliability of at least *r*= .65. Our initial power analysis showed that to achieve sufficient precision around the estimate of reliability, we would require 300 participants. Out of the 300 participants, we originally specified that 80 would be left-handers. This would give us power to apply our planned model-fitting analysis to left-handers as a separate group.  In practice, our final sample exceeded 300, because we discarded participants from individual tasks for failing to meet prespecified criteria (see below).

In total, 441 native English speakers were initially recruited through Prolific Academic (www.prolific.co) with all these participants completing both sessions. Prolific is a platform for recruiting and screening paid research participants according to experimenter-specified criteria. Participants were 18- to 50-years-old, had normal or corrected-to-normal vision and reported no history of neurological disease (e.g. head injury, brain tumour, stroke) or significant hearing loss or impairment.

@@

The basic demographics questionnaire asked participants to report age, gender, years in education, and whether they were bilingual. We also asked questions about handedness and footedness. For handedness, we asked: which is your dominant hand (i.e. which hand do you prefer to use for tasks such as writing, cutting, and catching a ball?). For footedness, we asked: which foot do you normally use to step up on a ladder/step? For both questions, participants could answer "left", "right", or "no preference / Don't know". No specific predictions were made about footedness, but it was included to provide a comprehensive picture of behavioural laterality.

@@

Adapted versions of the Miles test (Miles, 1929) and the Porta test (Porac & Coren, 1976) were used to determine each participant’s eye dominance in central gaze (i.e., when looking straight ahead). Each test classified participants as being either left or right eye dominant. We did not have specific predictions about ocular dominance but gathered this information to further characterise behavioural laterality of the sample. In practice, the two tests agreed perfectly, so we report here just the Miles test.

@@

The rhyme decision task was developed for this project as a novel visual half-field technique designed to engage lateralised language processes. On each trial, participants were asked to decide whether a foveally presented letter string rhymed with the name of an image shown in either the LVF or RVF. With online presentation, a particular concern is a potential for eye movements away from fixation. We relied on the fact that participants had to focus initially on the central written word to do the task, and thus the task encouraged fixation to a central location at the start of the trial. We subsequently displayed pictures under brief bilateral presentation, which has been shown to give adequate control for eye movements in previous visual half-field experiments with verbal naming (Beaumont, 1982). Studies that directly monitored eye movements in this setting reported failures of fixation on only 0.5% of trials (Geffen, Bradshaw & Nettleton, 1972; but see also Bourne, 2006). Pictures of familiar objects were presented for rhyme matching to the written word, to ensure that the participant would have to covertly name the picture, thereby engaging left hemisphere speech processing. We used covert rather than overt naming (cf. Hunter & Brysbaert, 2008) to make the task suitable for online presentation. This obviated the need for measuring latencies of oral responses, which would be challenging in the online context, both in terms of the need for a high-quality microphone, and possible intrusion of background noises interfering with recording. Furthermore, the evaluation of oral responses would require additional scoring following data collection which would be a taxing procedure for researchers and could introduce subjectivity into the measure. Voice recording would also require additional ethical approvals with regards to the storing and processing of the data.

@@

```{r read_dat, echo- FALSE, include= FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(corrr)
library(flextable)
library(cocor)
options(knitr.kable.NA = '') #should suppress NA in tables
# read in data from csv
#LIs <- read.csv('https://osf.io/x5jd2/download',stringsAsFactors = F)
#We will replace with read from OSF, but better stick with LIs to ensure final version is the one we use 
allsum <- read.csv('./data/allsum_out.csv',stringsAsFactors = F, na.strings = "NA")
allsum <- allsum[!is.na(allsum$origID), ]
```

```{r prepdat,echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}
# remove those without day 2 to make LIs file - we will just use this
# reformat NA
#identify missing cases for time 2
day2cols<-c('Day2_CF_acc_LI','Day2_DL_acc_LI','Day2.RDT_RT_LI','Day2_finger_LI')
allsum$day2.complete <- ifelse(is.na(allsum$day2.complete), 0, allsum$day2.complete)
LIs <- allsum[allsum$day2.complete == 1,]
